Inferring Fine-Scale Animal Behaviors using Hidden Markov Models

### 2020-06-08 Report/Sections/simulation.tex
Add plots to repo for overleaf

	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded). Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100*\lfloor Y_t/2 \rfloor + 1\}$. The parameters of the fine-scale behaviour Makov chain were set to be as follows:
	%
	$$\Gamma^{*(1)} = \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} \qquad 	\Gamma^{*(2)} = \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix}$$
	%
	where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$

@@

	\begin{align*}
	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.01 \right) & \\
	%
	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.01 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
	%
	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\\\
	%
	%
	a_{t,t^*}^{(n)} &\sim  \left\{\begin{array}{lr}
	-1 & w.p. \enspace 1/2 \\
	1  & w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) & \\
	%
	(b^{(n)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim  \left\{\begin{array}{lr}
	Gamma(1/n^2, 1) & \text{for } n \neq 2\\
	Gamma(100,1) & \text{for } n = 2
	\end{array}\right. & \\\\
	%
	%
	\hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
	%
	\hat{Y}^{*(n)}_{t,t^*}  &= -\hat{Y}^{*(100-n)}_{t,t^*}, & \qquad n \in \{51,\ldots,99\}
	\end{align*}

	Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete fourier transform of $\hat{Y}^*_{t,t^*}$:

	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$

	This gives a strong periodic component to acceleration when the subdive state $X^*_{t,t^*}= 2$. There are several practical reasons behind this construction:

	\begin{enumerate}
		\item $\hat{Y}^*_{t,t^*}$ is anti-symetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse fourier transform is real-valued.
		\item $\hat{Y}^{*(n)}_{t,t^*}$ decays like $1/n^2$ to facilitate continutity.
	\end{enumerate}

@@

Note that this process idoes not result in a continuous sequence $Y^*_t$ since the average values of consecutive 2-second segments jump. However, note that these average values are highly autocorrelated, so the jumps are not too severe. See figure (\ref{fig:sim_data}) for details. In addition, $Y^*_t$ has the following desirable properties:

\begin{enumerate}
	\item $Z^{*(1)}_{t,t^*+1} | Z^{*(1)}_{t,t^*} \sim \mathcal{N} \left(\mu = 0.9 Z^{*(1)}_{t,t^*} + 0.1, \sigma^2 = 0.01 \right)$
	\item $$Z^{*(2)}_{t,t^*} \sim \left\{\begin{array}{lr} 
	\text{Gamma}(\sum_{n=1}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 1 \\
	\text{Gamma}(20 + \sum_{n=2}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 2
	\end{array}\right. $$
\end{enumerate}

So we can directly compare the results of the CarHHMM with the ground truth.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=5in]{../../Plots/sim_data.png}
	\caption{Simulated Acceleration Data.}
	\label{fig:sim_data}
\end{figure}
I still need to put in the actually results from the simulation study- I had to rethink how to do the simulation in the first place since enforcing continuity was REALLY hard to do in a principled way.

>>>>

	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded). Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100*\lfloor Y_t/2 \rfloor + 1\}$. The parameters of the fine-scale behaviour Markov chain were set to be as follows:
	%
	$$\Gamma^{*(1)} = \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} \qquad 	\Gamma^{*(2)} = \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix}$$
	%
	where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$

@@

	\begin{align*}
    	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.0025 \right) & \\
    	%
    	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.0025 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
    	%
    	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\
	\end{align*}
	\begin{align*}
	    a_{t,t^*}^{(n)} &\sim  \left\{\begin{array}{lr}
    	-1 & w.p. \enspace 1/2 \\
    	1  & w.p. \enspace 1/2
    	\end{array}\right. \\
    	%
    	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) \\
    	%
    	(b^{(n)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
    	Gamma(1/n^2, 1) & \text{for } n \neq 2\\
    	Gamma(20,1) & \text{for } n = 2
    	\end{array}\right. 
	\end{align*}

    \begin{align*}
        \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
    	%
    	\hat{Y}^{*(n)}_{t,t^*}  &= -\hat{Y}^{*(100-n)}_{t,t^*}, & \qquad n \in \{51,\ldots,99\}
    \end{align*}

	\item Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$

@@

There are several practical reasons behind this construction. First, $\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(n)}_{t,t^*}$ also decays like $1/n^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window. Note, however, that $Y^*_{t,t^*}$ is not continuous in $t^*$ \textit{between} windows, as can be seen in figure (\ref{fig:fourier_example}). However, note the jumps are not too severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+100}$ are highly correlated. See figure (\ref{fig:sim_data}) for details.

\subsection{Ground-Truth parameters}

This generative model gives simulated data where the ground-truth of the CarHHMM is known. For the coarse scale, the ground truth is as follows:

\begin{align*}
    \delta &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix}\\
	\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
	% 
	Y_t|X_t = 1 &\sim Gamma\left(\mu = 15 s, \sigma = 5 s\right) \\
	Y_t|X_t = 2 &\sim Gamma\left(\mu = 60 s, \sigma = 10 s\right) \\
\end{align*}

For the fine-scale, we have:

\begin{align*}
    \delta^{*(1)} &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix} & \delta^{*(2)} &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix} \\
    %
    \Gamma^{*(1)} &= \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} & 	\Gamma^{*(2)} &= \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix} \\
\end{align*}

\begin{align*}
    \left(Z^{*(1)}_{t,t^*+100} | Z^{*(1)}_{t,t^*} \right) &\sim \mathcal{N} \left(\phi^* * Z^{*(1)}_{t,t^*} + \phi^* * \mu^*, \sigma^{*2} \right), & \mu^* = 1, \enspace \phi^* = 0.9, \enspace \sigma^{*2} = 0.025 \\
    %
    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) &\sim Gamma\left(\alpha^{*(1)}, \beta^* \right), & \alpha^{*(1)} = \sum_{n=1}^{\tilde{f}} \frac{1}{n^2}, \enspace \beta^* = 1 \\
    %
    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) &\sim Gamma\left(\alpha^{*(2)}, \beta^* \right), & \alpha^{*(2)} = 20 + \sum_{n=2}^{\tilde{f}} \frac{1}{n^2}, \enspace \beta^* = 1 \\
\end{align*}
%
The emission distribution of the 2-second average acceleration $Z_{t,t^*}^{*(1)}$ in particular can be reparameterized as an OU process:
%
$$dZ^{*(1)}_{t,t^*} = \beta^*(\gamma^* - Z^{*(1)}_{t,t^*})dt^* + \omega^* dW_{t,t^*}$$

$$\gamma^* = \mu^* = 1, \qquad \beta^* = -\frac{\log(\phi^*)}{\Delta t^*} = 0.0229, \qquad \omega^{*2} = \frac{2\sigma^{*2}\log(\phi)}{\Delta t (\phi^{*2}-1)} = 0.006$$

Note that the two fine-scale states $X^*_{t,t^*} = 1$ and $X^*_{t,t^*} = 2$ are identical except for the fact that the distribution of $Z^{*(2)}_{t,t^*}$ is shifted up in its shape parameter in fine-scale state 2. This corresponds to more periodic behavior in acceleration, but \textit{not} a higher mean acceleration.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=5in]{../Plots/sim_data.png}
	\caption{Simulated Acceleration Data.}
	\label{fig:sim_data}
\end{figure}

\subsection{Simulation Results}

Four different models were fit to the generated data shown in figure (\ref{fig:sim_data}):

\begin{enumerate}
    \item A CarHMM with no hierarchical component (CarHMM) using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations
    \item A Hierarchical hidden Markov model with no auto-correlation (HHMM) using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations.
    \item A CarHHMM using $Z^{(1)}$, but not $Z^{(2)}$, as fine-scale observations.
    \item A CarHHMM using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations
\end{enumerate}

Table (\ref{table:emis_dist}) gives the MLE estimates of the parameters for each model above.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
